Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2015-12-17 special-issue-ci/Response to reviewers-r2.tex

All acronyms have been checked and defined in the first occurrence,
some unnecessary ones (like JS for JavaScript) eliminated.

\begin{quote}
\textbf{
2) NodIO-W2 is not presented in the abstract.}
\end{quote}

We have now made a small introduction to it.

\begin{quote}
\textbf{3) Check the use of system vs framework.
We recommend to use only one of these two words all along the text.}
\end{quote}

@@

In general, we have opted for framework, since it is a client-server
framework and it is more precise than {\em system}

\begin{quote}
\textbf{4) Considering only one subsection inside a section makes harder the paper comprehensibility.
For instance, in case of section IV, we recommend removing the caption
related to subsection A and keeping everything in just one section.}
\end{quote}

@@

\begin{quote}
\textbf{9) The following recent reference is closely related to your paper and it should be cited:
J. Alcala-Fdez and Jose M. Alonso, A survey of fuzzy systems software: Taxonomy, current research trends and prospects, IEEE Transactions on Fuzzy Systems, 2015. Doi:10.1109/TFUZZ.2015.2426212}
\end{quote}

Added.

%-----------------------------------------------------------------------------
\section{Response to Reviewer \#1}

\begin{quote}
\textbf{I'm not, however, fully convinced about that the new experiment of solving Rastrigin's function be representative of a real world problem. The presented work, as is, has interesting merits, but the question of its usefulness remains uncertain if it would be used to solve a non-academic optimization problem. As the authors point out, reasonable performance is not ensured and their proposal could result in a pointless academic exercise.\\
After reading the revised paper, an arising question is why the
authors have used in Section IV a problem that is solved in around 1
second. An approach to simulate a hard problem could be to
artificially increase the evaluation time (e.g, by adding an idle
loop); this way, the problem would remain the same and the computing
time could be adjusted to try to gain some speed-up with the proposed
platform.\\}\end{quote}

@@

This is an useful suggestion. We will try to add that kind of
experiment in the future. In fact, some of our previous papers have
used a similar technique successfully.

\begin{quote}
\textbf{Minor comments:
- The paper structure should be enhanced: there are subsections IV.A and V.A but there are not
IV.B nor V.B.}\end{quote}

@@

As a response to editor's requests, this has been done.

\begin{quote}
\textbf{- In the first paragraph of Section V is said: ``! to perform the fifty runs'', but none is said before
about those fifty runs.}\end{quote}

@@

It is mentioned now at the beginning of the paragraph

\begin{quote}
\textbf{- Figures should be improved: Fig.2 has huge fonts (and the caption should include the time unit),
while others Fig.3, Fig.4, etc include very small fonts.}\end{quote}

@@

\begin{quote}
\textbf{- The second version of the systems is named NodIO-W2 in Section V, but it is referred as to
NodIO2 in the conclusions.}
\end{quote}

It is called the same way now. Thanks a lot for your suggestions, and
we expect these changes satisfactory.

%-----------------------------------------------------------------------------
\section{Response to Reviewer \#2}

\begin{quote}
\textbf{1.- The section discussing the threats to validity of the
  experiments merely discuss the security issues of the platform but
  not the threats to the validity of the own experimentation performed
  in the paper. I strongly recommend to authors to follow the
  recommendations and guidelines described in [1] to address such
  comment.\\}
\end{quote}

A comment to the threats to validity has been now included in the
paragraph next to the conclusion.

\begin{quote}
\textbf{2.- The figures 3,4 and 5 of the paper have not been modified according to our original comments: ``Figures should be readable even in gray-scale, so do not use red and blue to denote different elements, use a different shade or texture. Additionally, the axis labels and legends in most of the figures are unreadable, please increase its font size.'' I strongly recommend the authors to modify the figures to improve readability.}
\end{quote}

Previously, we have checked it in grayscale and thought it was OK. We
have done now as instructed.

We would also like to thank this reviewers for his work and helpful
comments.

### 2016-01-05 special-issue-ci/volunteer-clouds-ieee-ci-si-tr.tex
Separating in two parts
In a different branch, to keep history.

\documentclass[journal,onecolumn]{IEEEtran}

@@

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  \graphicspath{{../img/}}
\else
  \usepackage[dvips]{graphicx}
  \graphicspath{{../img/}}
\fi

@@

\title{Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources 
for Low Cost Distributed Evolutionary Computation }
% The word massive in the title is giving us problems.
% Why not change it to something about sharing resources?.
% Also lets not forget about volunteers, they can be a resource too
% and they cannot be replaced by any super computer

%  I propose something like:
% A Web Based Architecture to Harness Volunteer Computational Resources 
% for Low Cost Distributed Evolutionary Computation

% Some keywords missing: Cloud, Asynchronous, Pool-Based, Javascript-Based
% Open Source software.
% Lets not forget this issue is about software.

\author{Juan-J.~Merelo*$^1$, Mario Garc\'ia-Valdez$^2$, Pedro A. Castillo$^1$, Pablo Garc\'ia-S\'anchez$^1$, P. de las Cuevas*$^1$, Nuria Rico$^3$
\thanks{Manuscript submitted for review on \today.}%
\thanks{$^1$Dept. of Computer Architecture and Technology and CITIC University of Granada}%
\thanks{$^2$Dept. of Graduate Studies at Instituto Tecnol\'ogico de Tijuana}%
\thanks{$^3$Dept. of Statistics and Operational Research University of Granada}%
\thanks{E-mail addresses: {\tt jmerelo@ugr.es}, {\tt
    mario@tectijuana.edu.mx}, {\tt pacv@ugr.es}, {\tt pablogarcia@ugr.es},{\tt nrico@ugr.es}}%
\thanks{*Corresponding author.}%
}

@@

\begin{abstract}
Volunteer computing is extensively used in science and industry, but
it still has a long way to go in order to become a mainstream
technology in the same way the cloud technologies in general are. %I have changed "have" to "are"
The lack of general purpose and open
source tools with built-in analytics could be one of the reasons of
why this is so, as well as the lack of proofs of concept.
% (Paloma) The following sentence is way too long, and the "which" after "Javascript" is confusing, so one does not know if the which refers to the framework or Javascript by reading the whole sentence.
% I propose: "That is why in this paper we present NodIO, a framework used to create client-server architectures for distributed evolutionary algorithms, entirely written in JavaScript. Our framework needs minimal server support to work and is able to support persistent, asynchronous, distributed evolutionary algorithms running in the browser, as well as any other client consuming its REST application programming interface."
That is why
in this paper we present {\sf NodIO}, a framework used to create
client-server architectures for distributed evolutionary algorithms,
entirely written in JavaScript, which needs minimal server support to
work and that is able to support persistent, asynchronous, distributed
evolutionary algorithms running in the browser, as well as any other
client consuming its REST application programming interface. An
enhanced implementation, which we have called {\sf NodIO-W$^2$}, is
also presented; this version leverages the HTML5 Web Workers
specification to execute several clients per browser and do it in the
background, so that the the number of generations per user could be
improved. The architecture, along with some initial results on the
expected performance and scaling behavior when the number of users is
increased is introduced. % (Paloma) I've changed this "presented" for a synonym
We also use the algorithm data to find out
what kind of patterns the user behavior follows by examining the
resources they devote to our experiment, and try to fit them in % (Paloma) Changed. Phrasal verb was wrong.
 some
statistical distribution to create the bases of a model of user
behavior in volunteer computing experiments of this kind that can be
used by researchers to predict their performance. %(Paloma) Changed to "their" because of "experiments"
In general, our
experiments prove, first, that {\sf NodIO} is a flexible and
low-overhead platform to perform distributed computation experiments
that include volunteers, and then that user behavior follows some
patterns, and, thus, are amenable to a model.
\end{abstract}

@@

\begin{IEEEkeywords}
Evolutionary algorithms, evolutionary computation, distributed computing, internet computing,
social networks, volunteer computing, metacomputing, performance evaluation.
\end{IEEEkeywords}

@@

\IEEEPARstart{N}{owadays} there is a big and increasing number of
distributed evolutionary computation software frameworks available on
the market, usually as open source frameworks \cite{Parejo12Survey}. 
Most of these frameworks have 
two features: they are {\em desktop} applications, that is, they must
be compiled or run directly on the operating system layer and, second,
they assume that the nodes that are going to run the system are under
the control of developers or they at least have access to them. In
fact, these features are commonly found in most soft computing
frameworks \cite{7094263}.

There is, then, space for frameworks that do not make any of these
assumptions, such as the one presented in this paper, which will be
 named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can be embedded in any web-enabled device, including many over which 
developers have no control.

Some time ago, our group faced the problem of diminishing funds for
buying new hardware. This was aggravated by the increasing maintenance costs and
extended downtime resulting from the continuous failures of existing
clusters.
Considering this, we leveraged our experience in the design of web
applications since their inception, including JavaScript, and other
volunteer and
unconventional distributed evolutionary computing systems to 
design and release a new free framework that would allow anyone to
create a volunteer distributed evolutionary computation (EC) experiment using cloud resources as
servers, and browsers as clients.
% (Paloma) I think you should rewrite this whole sentence.
% For starters, focus in that you "leveraged" the problem through your deeply knowledge of... this and that. And "Thus/Then/Therefore", you decided... 
% (Mario) We leverage our experience in blah blah to design
% (Paloma) It may be... guess I was thinking in the noun, not in the
% transitive verb
% Please check if there's a problem here too.

%PABLO: He añadido el siguiente párrafo para el issue #31
There exist other volunteer computing frameworks written in compiled languages,
such as C++ or Java, that, in principle, can achieve better performance than
JavaScript \cite{2015arXiv151101088M}. However, one of the advantages of our architecture is that
clients do not need to install anything in their machines, they just
need to open a web page  
% Somewhat redundant 
% Javascript is the only language that can be executed in this way without any installation, 
% also removed the default, redundant with next sentence
in a compatible browser, such as Chrome, Safari or Firefox, which are
available in almost every desktop or mobile device. Other advantage is
that the source code can be directly examined by volunteers, because it is not
compiled as other options such as Java Applets, ActionScript or
ActiveX.
Also, as we will discuss in this paper, it can provide some extra
features such as enhanced security or operating system independence.

%PABLO: hay que enlazar con el siguiente párrafo
%Desktop clients have the same issues as any 
%volunteer computing system, including our % (Paloma) their?
% own: these issues are
%related to security as well as
%safety, which introduce

When developing volunteer based frameworks, designers must
also consider the whole {\em social} aspect in the design, with
issues related to security, trust, and privacy among others. 
The computing system becomes then a {\em techno-social system} \cite{vespignani2009predicting}.  
% (Paloma) Is the social aspect related to the word "safety"? It's not completely clear.
%a {\em social} aspect to the design of a
%computing system that makes it a {\em techno-social system}
In this paper, our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page running the experiment
determine its performance and how it can be modeled and, eventually,
predicted.

The rest of the paper is organized as follows: Next we present the
state of the art (Section \ref{sec:soa}) in web-based distributed
computational systems along with attempts to predict and model its behavior. 
Then, a detailed description of the proposed framework is presented
in Section \ref{sec:description}, followed by the baseline experiments
in Section \ref{sec:experiments}, and experiments using a new client
architecture that uses web workers in Section \ref{sec:w2}.
Finally, conclusions and future lines of work are presented in Section
\ref{sec:conclusion}.

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper % (Paloma) No really need for that "also", plus it complicates reading
establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, volunteer's safety, anonymity, hierarchy, ease of use, and
reasonable performance. But hierarchy, this is, the capability of
preferentially using resources in the same organization, is a feature
that could be either disregarded or included within adaptivity, which
leaves us with eight desirable properties. 
% (Paloma) Re-write. For example, "But is hierarchy, this is, the
% capability of preferentially using resources in the same
% organization, the one that could be either disregarded or included
% in adaptivity, which leaves us with eight desirable properties." 
% (Paloma) And I re-wrote what I proposed yesterday.
% thanks :-) -JJ

The fulfillment of these properties could be achieved by using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first. 
% (Paloma) These should be separate sentences
However, embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed by using desktop applications. 
Baratloo et al. tried in 1996 to create a framework with
the aforementioned qualities called {\em Charlotte}
\cite{baratloo1996charlotte}. This framework, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing libraries PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system, and access to a common data store in the
server. Additionally, since the Java virtual machine is a sandbox, it was safe to the user
and as it runs compiled bytecode it also guaranteed a certain level of
performance. % (Paloma) I've changed a bit this sentence to unify the use of past tense when talking about Charlotte
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. This
paper was later extended and published in a journal in 1999
\cite{baratloo1999charlotte}, but it is interesting to note that,
even in the early stages of the web, the possibility of
% (Paloma) Re-write. For example, "but it is interesting to note that, even in the early stages of the web, the possibility of..."
browser-based distributed computing was already researched.

Other systems, like JET, proposed by Soares et
al. \cite{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, enabling its use for science, %(Mario) Was easier to change "allowing"
since it allows % (Paloma) Use synonyms, like "yield" or "permit" 
the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\cite{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. SuperWeb is focused,
% (Paloma) Careful with starting using plural ("Other proposals", "are") and continuig with singular. If you are referring to "SuperWeb", change "It" for "SuperWeb"
however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers, mentioning
SPARCstations and PCs with Windows-95. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \cite{sarmenta-bayanihan} as well as the potential of
user misconduct when having access to the source code \cite{sarmenta-sabotagetolerance}. 
However, these early efforts by
Sarmenta once again used Java and not JavaScript, making them % (Paloma) making them
less widespread.

The main problem with Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; 
for instance numerous vulnerabilities have been discovered in 
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript \cite{flanagan2006javascript}, however, has become a set of standards
\cite{ECMA-262} that include the language and its components. Since
early on, it was adopted by the industry and also by scientists,
who used them for creating an evolutionary algorithm (EA from now on) on
the browser as early as 1998 \cite{jj-ppsn98}.

JavaScript can be used either for unwitting
\cite{unwitting-ec} or volunteer 
\cite{langdon:2005:metas,gecco07:workshop:dcor} distributed
evolutionary computation and it has been used ever since by several
authors, including more recent efforts
\cite{Desell:2008:AHG:1389095.1389273,duda2013distributed,DBLP:journals/corr/abs-0801-1210} 
that even
used the client's GPU \cite{duda2013gpu} or create an object-oriented
framework for evolution in the browser \cite{EvoStar2014:jsEO}. Many other researchers have
used Java \cite{chong:1999:jDGPi} and others have gone away from the
server-based paradigm to embrace peer to peer systems
\cite{jin2006constructing,10.1109/ICICSE.2008.99,DBLP:conf/3pgcic/GuervosMFEL12}. These computing
platforms avoid single points of failure (the server) but, since they
need a certain amount of infrastructure installed to start, the
threshold to join them is much lower. % (Paloma) Do yo mean... higher? A low threshold means that is easy to happen something.

Recent works propose the use of cloud computing services for the components of
client-server distributed EAs. Cloud resources
can reduce operational costs by outsourcing both hardware and software maintenance
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what
is available in most laboratories, allowing researchers to
scale algorithms at reasonable costs. Researchers have proposed the use of
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards},
and DropBox~\cite{mericloud}. The use of JavaScript to implement
server side components
gives software developers certain advantages in the cloud
paradigm, since it is a first-class citizen
in most Platforms as a Service (PaaS) products as the {\tt Node.js} framework is supported by most of them \cite{wood13:nodejs:paas}. Many web services
have native JavaScript libraries too, and also use the JSON (JavaScript Object Notation) format for data interchange. Another advantage is that
developers can use the same software tools for both client and server
side development.

JavaScript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}.
According to \cite{swanresearch2015}, the benefit of using this kind of paradigm in
metaheuristics optimization is that it allows an improvement in communicability,
reproductibility, interoperability, automated assembly, knowledge engineering,
and efficiency.

In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library {\sf NodEO} \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
when writing code and leverages 
the expertise acquired
% (Paloma) Not sure if "acquired" should be written before "expertise"... sounds better to me, anyways
 in the library itself.

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}. This is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or by farming out evaluations to all
the nodes available. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or be included in a more general study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

On the other hand, measuring the performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. 
%But the raw material of           
%volunteer computing, % (Paloma) Whith regard to?  Maybe it was colloquial/analogy
But some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal, and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
framework itself.

As far as we know, this paper presents one of the few experiments using % (Paloma) Change to "which" in order to not repeating "that" all the time
computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. 
%The methodology used
% (Paloma) I would say "the used methodology"
The algorithms used, as well as the methodology 
for gathering resources will be described next, 
together with the results obtained in this initial setup.
%---------------------------------------------------------------
\section{Description of the framework}
\label{sec:description}

@@

%---------------------------------------------------------------
\section{Baseline Experiments} %Cambié el nombre
\label{sec:experiments}

@@

These two experiments establish a baseline result and show that
the time to success depends on the population size, with a bigger population
contributing to have more diversity and thus speeding up the solution \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which these runs could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the month of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible.
%
\begin{table*}
\caption{Characteristics of the volunteer computing initial runs. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rr}
\hline
Date & \# experiments & Different IPs \\
\hline
April 4th 4/4 & 57 & 191 \\
April 24th 4/24 &  231 & 559 \\
July 31th 7/31 & 97 & 179 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of {\tt
    PUT}s per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
4/4 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 &  5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of IPs
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. Thus, the user can decide if to 
continue sharing resources or not. % (Paloma) Feature or disadvantage?
However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of gray indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
% ../data/time-vs-IPs-openshift.png 
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.png 
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade of gray). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next. % (Paloma) This sentences are sort of alone here... Could we extend a bit the first "this" to connect them to the previous paragraph ?

@@

After setting up baseline results for this particular problem, we will
present now results the results obtained. Next we will present the
details of the implementation and then the numeric results of the
experiment.

\subsubsection{Setup of Rastrigin's experiment in {\sf NodIO-W$^2$ }}

>>>>

\documentclass{article}

@@

  \usepackage[pdftex]{graphicx}
  \graphicspath{{../img/}}

@@

\title{}

\author{Juan-J.~Merelo*$^1$, Mario Garc\'ia-Valdez$^2$, Pedro
  A. Castillo$^1$,\\ 
Pablo Garc\'ia-S\'anchez$^1$, P. de las Cuevas*$^1$, Nuria Rico$^3$
\thanks{Manuscript submitted for review on \today.}%
\thanks{$^1$Dept. of Computer Architecture and Technology and CITIC University of Granada}%
\thanks{$^2$Dept. of Graduate Studies at Instituto Tecnol\'ogico de Tijuana}%
\thanks{$^3$Dept. of Statistics and Operational Research University of Granada}%
\thanks{E-mail addresses: {\tt jmerelo@ugr.es}, {\tt
    mario@tectijuana.edu.mx}, {\tt pacv@ugr.es}, {\tt pablogarcia@ugr.es},{\tt nrico@ugr.es}}%
\thanks{*Corresponding author.}%
}

@@

\begin{abstract}

\end{abstract}

@@

{\bf Keywords}
Evolutionary algorithms, evolutionary computation, distributed computing, internet computing,
social networks, volunteer computing, metacomputing, performance evaluation.

@@

The rest of the paper is organized as follows:

@@

%---------------------------------------------------------------
\section{Description of the framework}
\label{sec:description}

@@

%---------------------------------------------------------------
\section{Baseline Experiments} 
\label{sec:experiments}

@@

These two experiments establish a baseline result and show that
the time to success depends on the population size, with a bigger population
contributing to have more diversity and thus speeding up the solution \cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic.

@@
